The dynamics of starvation and recovery

### 2016-08-24 Manuscript/9.tex
Final analyses before first submission to PNAS

\begin{abstract} %246 words
  The eco-evolutionary dynamics of species is fundamentally linked to the
  energetic constraints of its constituent individuals.  Of particular
  importance are the tradeoffs between reproduction and the dynamics of
  starvation and recovery in resource-limited environments.  To elucidate the
  consequences of this tradeoff, we introduce a minimal nutritional
  state-structured model that incorporates two classes of consumer:
  nutritionally replete consumers that reproduce, and undernourished,
  non-reproducing consumers that are susceptible to mortality.  As a function
  of the transition rates between these replete and undernourished states
  that are determined by the presence or absence of resources, the consumer
  populations can either undergo cyclic dynamics or reach a steady state.  We
  obtain strong constraints on starvation and recovery rates by deriving
  allometric scaling relationships and find that population dynamics subject
  to these constraints can approach the cyclic regime but are typically
  driven to a steady state.  Moreover, we find that these rates fall within a
  `refuge' in parameter space, where the probability of extinction of the
  consumer population is minimized.  Thus we identify a potential mechanism
  that may both drive and constrain the dynamics of animal populations.  Our
  model provides a natural framework that predicts maximum body size for
  mammals by determining the relative stability of an otherwise homogeneous
  population to a mutant population with altered percent body fat.  For body
  masses $\lesssim 10^7$g, a more apidose invader dominates over the resident
  population, and vice versa for body mass $\gtrsim 10^7$g, thus providing a
  principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}
\keywords{foraging | starvation|reproduction}

@@

Several efforts have shown how a partitioning of $B$ between growth and
maintenance purposes can be used to derive a general equation for both the
growth trajectories and growth rates of organisms ranging from bacteria to
metazoans
\cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}. This relation is derived from the simple balance condition \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\dot{m}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
of the organism at any point in its development.  This balance has the
general solution \cite{bettencourt,Kempes:2012hy}
\begin{eqnarray}
\label{m1}
\left(\frac{m\left(t\right)}{M}\right)^{1-\eta}\!=1\!-\!\left[1\!-\!\left(\frac{m_{0}}{M}\right)^{1\!-\!\eta}\right]e^{-a\left(1\!-\!\eta\right)t/M^{1-\eta}}
\end{eqnarray}
where, for $\eta<1$, $M=(B_{0}/B_{m})^{1/(1-\eta)}$ is the asymptotic mass, $a=B_{0}/E_{m}$, and $m_0$ is mass at birth.  We now use this solution to define the timescale (see \cite{moses2008rmo} for a detailed presentation of these timescales) of reproduction and recovery from starvation. The time that it takes to reach a particular mass $\epsilon M$ is given by the timescale 
\begin{equation}
\label{t1}
\tau\left(\epsilon\right) = \ln\left[\frac{1-\left(m_{0}/M\right)^{1-\eta}}{1-\epsilon^{1-\eta}}\right]\frac{M^{1-\eta}}{a\left(1-\eta\right)}
\end{equation}
where we will define values of $\epsilon$ to describe a set of rates within our model. For the time to reproduce, $t_{\lambda}=\tau\left(\epsilon_{\lambda}\right)$, where $\epsilon_{\lambda}$ is the fraction of the asymptotic mass where an organism is reproductively mature and should be close to one (typically $\epsilon_{\lambda}\approx0.95$ \cite{West:2001bv}). The growth rate is then given by $\lambda=\ln\left(\upsilon\right)/t_{\lambda}$ where $\upsilon$ is the number of offspring produced, and for any constant value of $\epsilon_{\lambda}$ this will scale like $\lambda\propto M^{\eta-1}$ for $M>>m_{0}$ \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}.

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  Since
only certain tissues can be digested for energy (for example the brain cannot
be degraded to fuel metabolism), we define the rates for starvation, death,
and recovery by the timescales required to reach, or return from, specific
fractions of the replete-state mass (see the Supporting Information for details on these parameterizations).  We define
$m_{\sigma}=\epsilon_{\sigma} M$, where $\epsilon_{\sigma}<1$ is the fraction of
replete-state mass where reproduction ceases. This fraction will be modified
if tissue composition systematically scales with adult mass.  For example,
making use of the observation that body fat in mammals scales with overall
body size according to $M_{\rm fat}=f_{0}M^{\gamma}$ and assuming that once
this mass is fully digested the organism starves, this would imply that
$\epsilon_{\sigma}=1-f_{0}M^{\gamma}/M$. It follows that the
recovery timescale, $t_{\rho}$, is the time to go from
$m=\epsilon_{\sigma} \epsilon_{\lambda} M$ to $m=\epsilon_{\lambda}M$ (Figure \ref{fig:growth}). Using Eqs.~\eqref{m1} and \eqref{t1} this timescale is given by simply considering an adjusted starting mass of $m_{0}^{\prime}=\epsilon_{\sigma}\epsilon_{\lambda}M$, in which case
\begin{eqnarray}
t_{\rho}=\ln\left[\frac{1-\left(\epsilon_{\sigma}\epsilon_{\lambda}\right)^{1-\eta}}{1-\epsilon^{1-\eta}}\right]\frac{M^{1-\eta}}{a^{\prime}\left(1-\eta\right)}
\end{eqnarray}
where $a^{\prime}=B_{0}/E_{m}^{\prime}$ accounts for possible deviations in the biosynthetic energetics during recovery (see Supporting Information). It should be noted that more complicated ontogenetic models explicitly handle
storage \cite{hou}, whereas this feature is implicitly covered by the body
fat scaling in our framework.

@@

It should be noted that previous studies have used allometric constraints to explain the periodicity of cyclic populations
\cite{CalderIII:1983jd,Peterson:1984hj,Krukonis:1991fk}, suggesting a period
$\propto M^{0.25}$.  However this relation seems to hold only for some
species \cite{Hendriks:2012fc}, and potential drivers range from predator
and/or prey lifespans to competitive
dynamics~\cite{Kendall:1999iy,Hogstedt:2005cr}.
%It is more difficult to ascertain whether a population is truly cycling according to an internal or external dynamic~\cite{Kendall:1999iy} or if fluctuations are due to stochasticity, and this is particularly true for longer-lived organisms.
Statistically significant support for the existence of population cycles
among mammals is predominantly based on time series for small
mammals~\cite{Kendall:1998hl}, where our model would predict more pronounced transient dynamics, given how close these points are to
the Hopf bifurcation.  On the other hand, the longer gestational times and
the increased difficulty in measurements, precludes obtaining similar-quality
data for larger organisms.
\vspace{6mm}
%, we acknowledge that absence of evidence for cyclic dynamics among longer-lived species is not necessarily evidence of absence.
 %, emphasizing the role of starvation on dynamics.

@@

Furthermore, many previous studies have focused on the energy equivalence hypothesis which argues that the total energy use, $B_{tot}$, of a population is constant independent of species size (e.g. \cite{allen2002,enquist1998}). This hypothesis is based on observations showing that the abundance, $N$, of a species is proportional to the inverse of individual metabolism (e.g. $N\propto M^{-3/4}/B_{0}$)  (e.g. \cite{allen2002,enquist1998}). This is usually stated as $B_{tot}=NB=constant$ which has been shown to hold in mammalian and vascular plant communities (e.g. \cite{allen2002,enquist1998}). Figure \ref{} shows that both $F^{*}$ and $H^{*}$ scale like $M^{-\eta}$ over a wide range of organism sizes and Figure \ref{} shows that $F^{*}B$ is a constant over this same range. This result if remarkable because it illustrates that the steady state values of the NSM combined with the derived timescales naturally give rise to the energy equivalence result. Furthermore, the equivalence breaks down at distinct sizes at both the large and small end of mammals suggesting that these are hard limits on mammalian body sizes because organisms outside this range do not meet the constant efficiency obeyed by other populations. Significant deviations from constant energy use occur at $M \lesssim 1$ at the small end of mammals and $M\approx 6.5*10^7$ for the large end. Compellingly, this dynamic bound, which is determined by the rate of energetic recovery, is
close to the minimum observed mammalian body size of ca.\ 1.3-2.5 grams
(Fig.~\ref{fig:mass}b,c), a range that occurs as the recovery rate begins its
decline.  In addition to known transport limitations~\cite{West:2002ud}, we
suggest that an additional constraint of lower body size stems from the
dynamics of starvation.  This result mirrors other efforts \cite{Kempes:2012hy,Kempes:2016}
where at a given scale multiple limitations constrain the smallest possibilities for
life within a class of organisms.

@@

%Maximum mass bound
%Although there are upper bounds to the rate equations (e.g. when percent body fat becomes unity), they are not biologically feasible and we do not discuss them further.
A additional theoretical upper bound on mammalian body size is given by $\epsilon=0$, where mammals are entirely composed of metabolic reserves, and this occurs at a size of $M=8.3\times 10^8$, or $4.5\times$ the mass of a blue whale.
We determine a more realistic upper bound to body mass by assessing the
susceptibility of an otherwise homogeneous population to invasion by a
mutated subset of the population (denoted by ${}^\prime$) where individuals
have a modified proportion of body fat $M^\prime=M(1+\chi)$ where
$\chi \in [-0.5,0.5]$, thus altering the rates of starvation $\sigma$,
recovery $\rho$, and maintenance $\beta$.  There is no internal fixed point
that correspond to a state where both original residents and invaders coexist
(except for the trivial state $\chi=0$).  To assess the susceptibility to
invasion as a function of the invader mass, we determine which consumer has a
higher steady-state density for a given value of $\chi$.  We find that for
$1\leq M<10^6$g, having additional body fat ($\chi > 0$) results in a higher
steady-state invader population density
($H^{\prime *}+F^{\prime *}>H^*+F^*$).  Thus the invader has an intrinsic
advantage over the resident population.  However, for $M>10^6$, leaner
individuals ($\chi < 0$) have advantageous steady state densities.
%, and this is due to the changing
%covariance between energetic rates as a function of modified energetic
%reserves \sid{I don't understand the phrase after the comma AND STILL DON'T}.

The observed switch in susceptibility as a function of $\chi$ at
$M_{\rm opt}\approx10^6$ thus serves as an attractor, or an uninvadible evolutionary stable state, such that the NSM predicts organismal mass to increase if $M<M_{\rm opt}$ and
decrease if $M>M_{\rm opt}$.  Moreover, $M_{\rm opt}$, which is entirely
determined by the population-level consequences of energetic constraints, is
within an order of magnitude of the mass observed in the North American
mammalian fossil record~\cite{Alroy:1998p1594} and also the mass predicted
from an evolutionary model of body size evolution~\cite{Clauset:2009fh}.
While the state of the environment, as well as the competitive landscape,
will determine whether specific body sizes are selected for or
against~\cite{Saarinen:2014br}, we suggest that the starvation dynamics
outlined here may provide the driving mechanism for the evolution of larger
body size among terrestrial mammals.
%and we suggest that the starvation dynamics described here act in concert with these other factors is the only driver of body size evolution.

@@

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_Rates.pdf}
\caption{ 
Allometrically constrained starvation rate $\sigma$ (red) and recovery rate $\rho$ relative to the reproductive rate $\lambda$ (yellow) as a function of body mass.  
The rate of starvation is greater than the rate of reproduction for all realized terrestrial endotherm body sizes.
Mean values $\pm 20\%$ variation are shown by the shaded region for each rate.
}
\label{fig:gvs}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_FixedPoint.pdf}
\caption{ The transcritical (dashed) and Hopf bifurcation (solid) as a
  function of the starvation rate $\sigma$ and recovery rate $\rho$ for a 100 gram consumer.  These
  bifurcation conditions separate parameter space into infeasible, cyclic,
  and steady state dynamic regimes.  The color gradient shows the steady
  state densities for (\emph{A}) the resource $R^*$ and the (\emph{B}) energetically
  replete consumers $F^*$, (warmer colors denote higher densities).  Steady
  state densities for the energetically deficient consumers $H^*$ (not shown)
  scale with those for $F^*$.  }
\label{fig:fp}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_DataHopf.pdf}
\caption{ Transcritical (vertical lines) and Hopf bifurcations (curves) for
  allometrically determined starvation $\sigma$ and recovery $\rho$ rates as
  a function of different mammalian body sizes: $M=A\times10^1$ gram (blue) and
  $M=A\times10^6$ grams (green), where $A$ is a random uniform variable in $[1,9]$.
  }
\label{fig:hopf}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_ExtinctionAllometric.pdf}
\caption{ Probability of extinction for a 100g consumer as a function of the starvation rate $\sigma$ and recovery rate $\rho$, where the initial density is given as $A(F^*,H^*,R^*)$, with
  $A$ being a random uniform variable in $[0,2]$.  Extinction is defined as the
  population trajectory falling below $0.1~\times$ the allometrically
  constrained steady state. The white point denotes the allometrically constrained starvation and recovery rate.}
\label{fig:ext}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{fig_FPAllometric.pdf}
\caption{ (\emph{A}) Consumer steady states $F^*$ and $H^*$ as a function of
  body mass. 
  (\emph{B}) Resource steady state $R^*$ as a function of consumer body mass.
  (\emph{C}) Total energetic use $B_{\rm tot}$ of consumer populations at the steady state as a function of body mass.}
\label{fig:mass}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_Invasion.pdf}
\caption{ Invasion feasibility for organisms with a proportional change in
  mass $\chi$ against a population with a resident body mass $M$.  The blue
  region denotes proportions of modified mass $\chi$ resulting in successful invasion.  The
  red region denotes values of $\chi$ that result in a mass that is below the
  starvation threshold and is thus infeasible.
  Arrows point to the predicted optimal mass $M_{\rm opt}=8.43\times 10^6$, which serves as the evolutionary stable state.}
\label{fig:invasion}
\end{figure*}

>>>>

\begin{abstract} %246 words
  The eco-evolutionary dynamics of species is fundamentally linked to the
  energetic constraints of its constituent individuals.  Of particular
  importance are the tradeoffs between reproduction and the dynamics of
  starvation and recovery in resource-limited environments.  To elucidate the
  consequences of this tradeoff, we introduce a minimal nutritional
  state-structured model that incorporates two classes of consumer:
  nutritionally replete consumers that reproduce, and undernourished,
  non-reproducing consumers that are susceptible to mortality.  As a function
  of the transition rates between these replete and undernourished states
  that are determined by the presence or absence of resources, the consumer
  populations can either undergo cyclic dynamics or reach a steady state.  We
  obtain strong constraints on starvation and recovery rates by deriving
  allometric scaling relationships and find that population dynamics subject
  to these constraints can approach the cyclic regime but are typically
  driven to a steady state.  Moreover, we find these rates fall within a
  `refuge' in parameter space, where the probability of extinction of the
  consumer population is minimized.  Thus we identify a potential mechanism
  that may both drive and constrain the dynamics of animal populations.  Our
  model provides a natural framework that predicts maximum body size for
  mammals by determining the relative stability of an otherwise homogeneous
  population to a mutant population with altered percent body fat.  For body
  masses $\lesssim 10^6$g, a more apidose invader dominates over the resident
  population, and vice versa for body mass $\gtrsim 10^6$g, thus providing a
  principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}
\keywords{foraging | starvation|reproduction}

@@

Several efforts have shown how a partitioning of $B$ between growth and
maintenance purposes can be used to derive a general equation for both the
growth trajectories and growth rates of organisms ranging from bacteria to
metazoans
\cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}. This relation is derived from the simple balance condition \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\dot{m}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
of the organism at any point in its development.  This balance has the
general solution \cite{bettencourt,Kempes:2012hy}
\begin{eqnarray}
\label{m1}
\left(\frac{m\left(t\right)}{M}\right)^{1-\eta}\!=1\!-\!\left[1\!-\!\left(\frac{m_{0}}{M}\right)^{1\!-\!\eta}\right]e^{-a\left(1\!-\!\eta\right)t/M^{1-\eta}}
\end{eqnarray}
where, for $\eta<1$, $M=(B_{0}/B_{m})^{1/(1-\eta)}$ is the asymptotic mass, $a=B_{0}/E_{m}$, and $m_0$ is mass at birth.  We now use this solution to define the timescale (see \cite{moses2008rmo} for a detailed presentation of these timescales) of reproduction and recovery from starvation. The time that it takes to reach a particular mass $\epsilon M$ is given by the timescale 
\begin{equation}
\label{t1}
\tau\left(\epsilon\right) = \ln\left[\frac{1-\left(m_{0}/M\right)^{1-\eta}}{1-\epsilon^{1-\eta}}\right]\frac{M^{1-\eta}}{a\left(1-\eta\right)}
\end{equation}
where we will define values of $\epsilon$ to describe a set of rates within our model. For the time to reproduce, $t_{\lambda}=\tau\left(\epsilon_{\lambda}\right)$, where $\epsilon_{\lambda}$ is the fraction of the asymptotic mass where an organism is reproductively mature and should be close to one (typically $\epsilon_{\lambda}\approx0.95$ \cite{West:2001bv}). The growth rate is then given by $\lambda=\ln\left(\upsilon\right)/t_{\lambda}$ where $\upsilon$ is the number of offspring produced (see supplement), and for any constant value of $\epsilon_{\lambda}$ this will scale like $\lambda\propto M^{\eta-1}$ for $M>>m_{0}$ \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}.

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  Since
only certain tissues can be digested for energy (for example the brain cannot
be degraded to fuel metabolism), we define the rates for starvation, death,
and recovery by the timescales required to reach, or return from, specific
fractions of the replete-state mass (see the Supporting Information for details on these parameterizations).  We define
$m_{\sigma}=\epsilon_{\sigma} M$, where $\epsilon_{\sigma}<1$ is the fraction of
replete-state mass where reproduction ceases. This fraction will be modified
if tissue composition systematically scales with adult mass.  For example,
making use of the observation that body fat in mammals scales with overall
body size according to $M_{\rm fat}=f_{0}M^{\gamma}$ and assuming that once
this mass is fully digested the organism starves, this would imply that
$\epsilon_{\sigma}=1-f_{0}M^{\gamma}/M$. It follows that the
recovery timescale, $t_{\rho}$, is the time to go from
$m=\epsilon_{\sigma} \epsilon_{\lambda} M$ to $m=\epsilon_{\lambda}M$. Using Eqs.~\eqref{m1} and \eqref{t1} this timescale is given by simply considering an adjusted starting mass of $m_{0}^{\prime}=\epsilon_{\sigma}\epsilon_{\lambda}M$, in which case
\begin{eqnarray}
t_{\rho}=\ln\left[\frac{1-\left(\epsilon_{\sigma}\epsilon_{\lambda}\right)^{1-\eta}}{1-\epsilon^{1-\eta}}\right]\frac{M^{1-\eta}}{a^{\prime}\left(1-\eta\right)}
\end{eqnarray}
where $a^{\prime}$ accounts for possible deviations in the biosynthetic energetics during recovery (see Supporting Information). It should be noted that more complicated ontogenetic models explicitly handle
storage \cite{hou}, whereas this feature is implicitly covered by the body
fat scaling in our framework.

@@

\sid{I don't see the point of the paragraph below.}
Previous studies have used allometric constraints to explain the periodicity
of cyclic populations
\cite{CalderIII:1983jd,Peterson:1984hj,Krukonis:1991fk}, suggesting a period
$\propto M^{0.25}$.  However this relation seems to hold only for some
species \cite{Hendriks:2012fc}, and potential drivers range from predator
and/or prey lifespans to competitive
dynamics~\cite{Kendall:1999iy,Hogstedt:2005cr}.
%It is more difficult to ascertain whether a population is truly cycling according to an internal or external dynamic~\cite{Kendall:1999iy} or if fluctuations are due to stochasticity, and this is particularly true for longer-lived organisms.
Statistically significant support for the existence of population cycles
among mammals is predominantly based on time series for small
mammals~\cite{Kendall:1998hl}, where our model would predict more pronounced transient dynamics, given how close these points are to
the Hopf bifurcation.  On the other hand, the longer gestational times and
the increased difficulty in measurements, precludes obtaining similar-quality
data for larger organisms.
\vspace{6mm}
%, we acknowledge that absence of evidence for cyclic dynamics among longer-lived species is not necessarily evidence of absence.
 %, emphasizing the role of starvation on dynamics.

@@

Furthermore, many previous studies have focused on the energy equivalence hypothesis which argues that the total energy use, $B_{tot}$, of a population is constant independent of species size (e.g. \cite{allen2002,enquist1998}). This hypothesis is based on observations showing that the abundance, $N$, of a species is proportional to the inverse of individual metabolism (e.g. $N\propto M^{-3/4}/B_{0}$)  (e.g. \cite{allen2002,enquist1998}). This is usually stated as $B_{tot}=NB=constant$ which has been shown to hold in mammalian and vascular plant communities (e.g. \cite{allen2002,enquist1998}). Figure \ref{} shows that both $F^{*}$ and $H^{*}$ scale like $M^{-\eta}$ over a wide range of organism sizes and Figure \ref{} shows that $F^{*}B$ is a constant over this same range. This result if remarkable because it illustrates that the steady state values of the NSM combined with the derived timescales naturally give rise to the energy equivalence result. Furthermore, the equivalence breaks down at distinct sizes at both the large and small end of mammals suggesting that these are hard limits on mammalian body sizes because organisms outside this range do not meet the constant efficiency obeyed by other populations. Significant deviations from constant energy use occur at $Justin -VALUE$ at the small end of mammals and $Justin -VALUE$ for the large end. Compellingly, this dynamic bound, which is determined by the rate of energetic recovery, is
close to the minimum observed mammalian body size of ca.\ 1.3-2.5 grams
(Fig.~\ref{fig:mass}b,c), a range that occurs as the recovery rate begins its
decline.  In addition to known transport limitations~\cite{West:2002ud}, we
suggest that an additional constraint of lower body size stems from the
dynamics of starvation.  This result mirrors other efforts \cite{Kempes:2012hy,Kempes:2016}
where at a given scale multiple limitations constrain the smallest possibilities for
life within a class of organisms.

@@

%Maximum mass bound
%Although there are upper bounds to the rate equations (e.g. when percent body fat becomes unity), they are not biologically feasible and we do not discuss them further.
A additional theoretical upper bound on mammalian body size is given by $\epsilon=0$, where mammals are entirely composed of metabolic reserves, and this occurs at a size of $M=8.3\times 10^8$, or $4.5\times$ the mass of a blue whale.
We determine a more realistic upper bound to body mass by assessing the
susceptibility of an otherwise homogeneous population to invasion by a
mutated subset of the population (denoted by ${}^\prime$) where individuals
have a modified proportion of body fat $M^\prime=M(1+\chi)$ where
$\chi \in [-0.5,0.5]$, thus altering the rates of starvation $\sigma$,
recovery $\rho$, and maintenance $\beta$.  There is no internal fixed point
that correspond to a state where both original residents and invaders coexist
(except for the trivial state $\chi=0$).  To assess the susceptibility to
invasion as a function of the invader mass, we determine which consumer has a
higher steady-state density for a given value of $\chi$.  We find that for
$1\leq M<10^6$g, having additional body fat ($\chi > 0$) results in a higher
steady-state invader population density
($H^{\prime *}+F^{\prime *}>H^*+F^*$).  Thus the invader has an intrinsic
advantage over the resident population.  However, for $M>10^6$, leaner
individuals ($\chi < 0$) have the advantage, and this is due to the changing
covariance between energetic rates as a function of modified energetic
reserves \sid{I don't understand the phrase after the comma AND STILL DON'T}.

The observed switch in susceptibility as a function of $\chi$ at
$M_{\rm opt}\approx10^6$ thus serves as an attractor, where over evolutionary
time the NSM predicts organismal mass to increase if $M<M_{\rm opt}$ and
decrease if $M>M_{\rm opt}$.  Moreover, $M_{\rm opt}$, which is entirely
determined by the population-level consequences of energetic constraints, is
within an order of magnitude of the mass observed in the North American
mammalian fossil record~\cite{Alroy:1998p1594} and also the mass predicted
from an evolutionary model of body size evolution~\cite{Clauset:2009fh}.
While the state of the environment, as well as the competitive landscape,
will determine whether specific body sizes are selected for or
against~\cite{Saarinen:2014br}, we suggest that the starvation dynamics
outlined here may provide the driving mechanism for the evolution of larger
body size among terrestrial mammals.
%and we suggest that the starvation dynamics described here act in concert with these other factors is the only driver of body size evolution.

@@

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_GvS.pdf}
\caption{ Allometrically constrained starvation rate $\sigma$ (red) vs.\
  reproductive rate $\lambda$ (yellow) as a function of mass $M$.  The rate
  of starvation is greater than the rate of reproduction for all realized
  terrestrial endotherm body sizes.  Each data point is determined by fixing
  the mass $M$ and then choosing parameters in a $\pm 20$\% range about the
  allometric values.}
\label{fig:gvs}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_FixedPoint.pdf}
\caption{ The transcritical (dashed) and Hopf bifurcation (solid) as a
  function of the starvation rate $\sigma$ and recovery rate $\rho$.  These
  bifurcation conditions separate parameter space into infeasible, cyclic,
  and steady state dynamic regimes.  The color gradient shows the steady
  state densities for (A) the resource $R^*$ and the (B) energetically
  replete consumers $F^*$, (warmer colors denote higher densities).  Steady
  state densities for the energetically deficient consumers $H^*$ (not shown)
  closely mirror those for $F^*$.  }
\label{fig:fp}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_DataHopf.pdf}
\caption{ Transcritical (vertical lines) and Hopf bifurcations (curves) with
  allometrically determined starvation $\sigma$ and recovery $\rho$ rates as
  a function of minimum and maximum mammalian body sizes: $1$ gram (blue) and
  $10^7$ grams (orange), respectively.  Replicates show the role of the 20\%
  variation about the mean allometric parameters, which influences both the
  energetic rates as well as the position of the TC and Hopf bifurcations.  }
\label{fig:hopf}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_ExtinctionAllometric.pdf}
\caption{ Probability of extinction for 1000 consumer population trajectories
  as a function of $\sigma/\rho$ for initial densities $A(F^*,H^*,R^*)$, with
  $A$ a random uniform variable in $[0,2]$.  Extinction is defined as the
  population trajectory going below $0.2~\times$ the allometrically
  constrained steady state at any time between $10^2$ and $\leq 10^6$.
  \sid{Insert a distribution instead of a scatter plot above the main panel
    and describe it precisely} The values above the extinction plot are the
  allometrically constrained $\sigma/\rho$ with 20\% variation around the
  mean.  }
\label{fig:ext}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.45\textwidth]{fig_FPAllometric.pdf}
\caption{ (\emph{A}) Consumer steady states $F^*$ and $H^*$ as a function of
  body size.  Energetic rates as a function of body size, with the ratio
  $\sigma/\rho$ (\emph{B}) and both $\sigma$ (red) and $\rho$ (purple;
  \emph{C}) drawn separately with 20\% variation around the mean.  Steady
  state densities decline sharply at $M=M_{\rm min}$ due to the precipitous
  decrease in the recovery rate.  The minimum mammalian body mass (Etruscan
  shrew) is denoted by the orange shaded region that coincides with the
  decline of the recovery rate.  }
\label{fig:mass}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.5\textwidth]{fig_Invasion.pdf}
\caption{ Invasion feasibility for organisms with a proportional change in
  mass $\chi$ against a population with a resident body mass $M$.  The blue
  region denotes values of $\chi$ that result in successful invasion.  The
  red region denotes values of $\chi$ that result in a mass that is below the
  starvation threshold and is thus infeasible.  }
\label{fig:invasion}
\end{figure*}
